Fast Sorting on Modern Computers

### 2005-07-19 docs/tech_report/radix.tex
fixed the key position in a graph rewrote some of the results in radixosrt

The instruction count of base radixsort, shown in Figure \ref{radix instructions} is
very low. It is lower than even quicksort for the smallest set, and it
flat-lines immediately (after amortisation of the loop overhead). The number of
cache misses in Figure \ref{radix level 2 misses} is exactly the same, until
the array no longer fits into the cache. The number of misses then leaps up to
2.5 per key, which is roughly what was predicted.

Cache radixsort performs as predicted aswell. By sorting keys between two
arrays instead of copying a partially sorted array back, the instruction count
has been reduced by more than 10\%, and the number of level 2 cache misses
have been more than halved, to just over one miss per key.

The number of branch misses, shown in Figure \ref{radix branch misses} was also
extremely low. Despite 10 branches per key, there was less than .08 of a branch
miss per key. The number of branches itself is quite low, as several of the
sorts considered had nearly that number of misses. However, it is not
unexpected. The number of branches reduces when copying back is removed, and
while this reduces the number of mispredictions, there are so few to begin with
that this result is not significant.

With the low number of branch misses, a much better overall performance is
expected than is achieved. When in the cache, it performs only slightly worse
than quicksort, but once the data set grows larger, the gap widens.  However,
the results, shown in Figure \ref{radix cycles}, indicate that radixsort will
never increase beyond 350 cycles per key, meaning that for larger sets it will
perform significantly better than quicksort. Cache radixsort never increases
beyond 300 cycles, though it has relatively more cycles when the arrays are
smaller.

The most curious result is that all simulations show that cache radixsort
performing better than the base radixsort, yet the Pentium 4 results show that
base radixsort performs better for smaller arrays.

LaMarca's radixsort results are quite different to those presented here. His
instruction count graph has a logarithmic shape: it begins very high and sharply
falls before flattening out. By contrast, the graph here is almost completely
flat. A similar thing happens in both cache miss graphs: a sharp dip in
LaMarca's results, and a flat-line with a slight slope in these results. Also,
in both cases, once the array no longer fits in the cache, the number of misses
jumps to around 2.5, where they stay.

The cycles per key in both cases are unsurprising based on instruction count and
cache miss results: LaMarca's have a sharp fall before rising slightly once the
data set no longer fits in the cache; the result presented here flat-line before rising
slightly for the same reason. While both results agree for large data sets, they do not
for smaller sets.

\section{Future Work}
\label{radix future work}
Combining other sorts with radixsort could also be possible. An optimization to
mergesort was to do a k-way merge at the end, but this dramatically increased
the instruction count. Using a type of radixsort here would be complicated, but
may be possible. Alternatively, sorting inside the cache using radixsort is far
faster than merging. The k-way merge could then be performed at the end. In
practice, though, radixsort is far faster than mergesort, has a lower
instruction count, and less cache misses. While these steps could improve
mergesort, they are more likely to slow radixsort, with no advantage.

Radixsort could also be faster than multi-quicksort's initial step. Replacing
this with a single radix step would split the array into 256 containers, each of
which could be quicksorted. Alternatively, a smaller radix, based on the size of
the array, could be generated at run time. Using LaMarca's analysis,
$C/3$ containers need to be generated. With $N$ as the number of keys in the
array, and $C$ is the number of keys which fit in the level 2 cache, the radix
can be calculated with the formula:
$$2^{\lceil{}log_2(\frac{3N}{CacheSize})\rceil{}}$$
This will only work for random, evenly distributed keys. If all the keys only
vary in their least significant bits, then this will have no effect.

>>>>

The instruction count of base radixsort, shown in Figure \ref{radix instructions} is
very low. It is lower than even quicksort for the smallest set, and it
flat-lines immediately (after amortisation of the loop overhead). The number of
cache misses in Figure \ref{radix level 2 misses} is exactly the same, until
the array no longer fits into the cache. The number of misses then leaps up to
2.5 per key, which is slightly more than was predicted. The extra is likely due
to the unpredictable way that the source array is accessed as the keys are
copied to the auxilliary array.

Cache radixsort performs as predicted aswell. By sorting keys between two
arrays instead of copying a partially sorted array back, the instruction count
has been reduced by more than 10\%, and the number of level 2 cache misses
have been more than halved, to one and a quarter miss per key. This is exactly
what is expected, due to there being nine passes over the array, and a one miss
per pass per cache line.

The number of misses while the array fits in the cache is also obvious: there
are two compulsory misses per key - one for each array - which divided by the
keys per cache line results in exactly one miss in four.

Aligned cache radixsort performs exactly the same as cache radixsort, indicating
that the alignment, and the potentially large memory overhead, are not
necessary.

The number of branch misses, shown in Figure \ref{radix branch misses} was also
extremely low. Despite 2 branches per key, there was less than .08 of a branch
miss per key. The number of branches itself is quite low, as several of the
sorts considered had nearly that number of misses. However, it is not
unexpected. The number of branches reduces when copying back is removed, and
while this reduces the number of mispredictions, there are so few to begin with
that this result is not significant.

With the low number of branch misses, a much better overall performance is
expected than is achieved. When in the cache, it performs only slightly worse
than quicksort, but once the data set grows larger, the gap widens.  However,
the results, shown in Figure \ref{radix cycles}, indicate that radixsort will
not increase much beyond 350 cycles per key, meaning that for larger sets it will
perform significantly better than quicksort. Cache radixsort does not reach 
250 cycles per key, though it has relatively more cycles when the arrays are
smaller.

LaMarca's radixsort results are quite different to those presented here. His
instruction count graph has a logarithmic shape: it begins very high and sharply
falls before flattening out. By contrast, the graph here is almost completely
flat. A similar thing happens in both cache miss graphs: a sharp dip in
LaMarca's results, and a flat-line with a slight slope in these results. Also,
in both cases, once the array no longer fits in the cache, the number of misses
jumps to around 2.5, where they stay. 
%TODO rewrite this bit a little
%TODO cache radixsort is faster than quicksort in all cases
The cycles per key in both cases are unsurprising based on instruction count and
cache miss results: LaMarca's have a sharp fall before rising slightly once the
data set no longer fits in the cache; the result presented here flat-line before rising
slightly for the same reason. While both results agree for large data sets, they do not
for smaller sets.

\section{Future Work}
\label{radix future work}
Combining other sorts with radixsort could be possible.  Radixsort should be be
faster than multi-quicksort's initial step. Replacing this with a single radix
step would split the array into 256 containers, each of which could be
quicksorted. Alternatively, a smaller radix, based on the size of the array,
could be generated at run time. Using LaMarca's analysis, $C/3$ containers need
to be generated. With $N$ as the number of keys in the array, and $C$ is the
number of keys which fit in the level 2 cache, the radix can be calculated with
the formula:
$$2^{\lceil{}log_2(\frac{3N}{CacheSize})\rceil{}}$$
This will only work for random, evenly distributed keys. If all the keys only
vary in their least significant bits, then this will have no effect.
